Machine Learning for Mechanistic Models of Metapopulation Dynamics

### 2023-11-01 ms.Rnw
tidying

\section{Introduction}
\label{sec:intro}

@@

\section{Review of metapopulation models and inference methods}
\label{sec:review}

@@

Population dynamics may be approximately linear on a log scale, and this has been used to develop linearization methods for epidemiological time series analysis \cite{bjornstad02} that have been extended to metapopulation analysis \cite{xia04}.
This provides a numerically convenient set of tools, but requires scientists to work within a limited class of models. The ensemble Kalman filter (EnKF) developed in the context of massive geophysical model, combines an ensemble representation with a computationally efficient update rule inspired by the scalable linear Kalman filter, providing an approach with excellent scalability \cite{evensen09book,evensen22}.
For biological systems, EnKF was first demonstrated as a computationally convenient tool for compartment models at a single location \cite{shaman12,yang14}.
Subsequently, it has been applied for epidemiological metapopulation inference \cite{li20,kramer20}.
However, the linearization in the EnKF filter update rule can be problematic for highly nonlinear systems \cite{evensen22}.
Further, a linear update rule is not appropriate for small, discrete populations unless EnKF is embedded within a MCMC algorithm \cite{katzfuss19}.

@@

\section{Metapopulation analysis of COVID-19 spread in China}
\label{sec:covid}
Models are an irreplaceable tool to inform public policy, despite delicate issues in their implementation and interpretation \cite{mccabe21,saltelli20}.
Some of the difficulties are operational, others are conceptual.
Operationally, we seek to fit complex models using statistically valid, reproducible and transparent methods.
Conceptual difficulties arise when drawing causal conclusions from fitting models to observational data, giving rise to opportunities for incorrect conclusions due to missing variables or other forms of model misspecification.
A strength of metapopulation modeling is that the model can be build upon established scientific knowledge, which may give some protection against gross forms of model misspecification (for example, asserting the existence of latent dynamic variables which simply do not exist).
A model assimilated to data guarantees that assumptions have been framed in a way consistent with certain facts, and evidence for predictive skill can support the value of the model construction.
Indeed, it can be practically impossible to make sense of the nonlinear stochastic interactions driving biological dynamics without representing them via a model \cite{mccabe21,lofgren14}.
We therefore seek to harness the power of models while assessing, acknowledging and minimizing their weaknesses.

As an example, we reconsider the influential analysis of COVID-19 from early in the pandemic by Li et al. \cite{li20}, that provided estimates of transmission parameters and the effect of lockdown using the limited data available at the time.
Other teams have fitted models to address similar questions \cite{kramer20-science,yang21,brett23} but the study by \cite{li20} is distinctive for fitting a stochastic mechanistic metapopulation model to extensive spatiotemporal data for 373 Chinese cities.
The results were published in May, 2020, based on reported cases from January 10 to February 8 of that year.
The state-of-the-art spatiotemporal analysis was possible on an urgent timescale because the team of researchers had developed their methodology in a sequence of previous situations \cite{shaman12,yang14,yang15,pei18}.
The paper is written to a high standard of reproducibility, and the main results are strengthened by various supporting analyses in an extensive supplement.
While examining the points mentioned above, we have identified various weaknesses that could have been mitigated by adhering to the aforementioned recommendations.
Our goal is not to criticize any specific paper, but rather to build on the timely analysis of \cite{li20} to demonstrate how recently developed techniques provide possibilities to carry out superior data analysis in future.

@@

We avoided the issue of model {\LiMobility} by adding some additional mobility based on a gravity movement model, as described in Supplementary Sec.~\suppSecMobility, giving rise to model~{\LiParams}.
We implemented an additional adjustment between models {\LiMobility} and {\LiParams} to align the measurement model with the ensemble Kalman filter (EnKF) inference method presented by \cite{li20}.
That EnKF implementation involved specifying a quanitity called the observation error variance, defined as a function of the observed cases, to quantify the uncertainty in the measurements.
Within the POMP specification, the measurement variance can depend on the latent state but not directly on the observed data, neverthless we modified the measurement model for~{\LiParams}to have equivalen scaling to the choice of \citep{li20}.

@@

All parameter estimates are reported in Table~\suppTableResults.
Here, we discuss the estimated basic reproductive number (i.e., the expected number of secondary infections from one index case in a fully susceptible population), denoted by $\Rzero^{\before}$ and $\Rzero^{\after}$ before and after the January 23rd lockdown.
Our estimates for model {\RevisedModelConstrained}, are $\Rzero^{\before}=\Sexpr{myround(ci_c["R0_be", "mle"],2)}$ with confidence interval (CI) (\Sexpr{myround(ci_c["R0_be", "ci1"],2)},\Sexpr{myround(ci_c["R0_be", "ci2"],2)}) and $\Rzero^{\after}=\Sexpr{myround(ci_c["R0_af","mle"],2)}$  with CI (\Sexpr{myround(ci_c["R0_af", "ci1"],2)},\Sexpr{myround(ci_c["R0_af","ci2"],2)}), where the estimates and their associated 95\% CIs obtained by profile likelihood (Supplementary Sec.~\suppSecInference).
This implies that the Chinese government non-pharmaceutical interventions instituted on and around January 23 reduced $\Rzero$ by a factor of $\Sexpr{myround(ci_c["R0_be","mle"]/ci_c["R0_af","mle"],1)}$.
By contrast, the estimates of \cite{li20} are $\widetilde\Rzero^{\before}=2.38$  with CI $(2.03,2.77)$ and $\widetilde\Rzero^{\after}=0.98$  with CI $(0.83,1.16)$, implying reduction by a factor of $\Sexpr{myround(2.38/0.98,1)}$.
For comparison, interventions implemented across a panel of 41 countries (34 European) were estimated to reduce $\Rzero$ by a factor of $\Sexpr{myround(1/(1-0.77),1)}$  with CI $(\Sexpr{myround(1/(1-0.65),1)},\Sexpr{myround(1/(1-0.85),1)})$ \citep{brauner21}.
Our estimate for $\Rzero$ before lockdown is toward the high end of comparable estimates, reviewed by \citep{park20-jcm}.
An alternative metaopopulation analysis of the pre-lockdown China data, with a deterministic transmission model, obtained an $\Rzero$ estimate of 3.11  with CI  (2.39,4.13) \citep{read21}.

Our model inherits the property of \citep{li20} that infections arising during the pre-lockdown period will generally be reported during the lockdown, due to the reporting delay modeled as a distributed delay with a mean of 9 days pre-lockdown and 6 days post-lockdown.
Thus, the model is permitted to explain the data by inferring rapid, unreported spread prior to January 23.
Despite this shared constraint on the form of the model, conclusions of our analysis differ from \cite{li20}.
Beyond the estimates of $\Rzero$, a notable difference is that we find the estimated transmissibility of observed cases is close to that of unobserved cases, especially before lockdown (Table~S1).

@@

\section{Discussion}
\label{sec:conclusion}

@@

Software environments are critical for providing data analysis that is not only reproducible but also readily extendable.
Scientists developing a data analysis should develop an environment that empowers them to explore their own models and data, and then they should share this environment as part of the publication process.
In practice, this involves encapsulating data analysis within a software package  that immerses the user in a documented environment where the models, methods and data used for the article can be readily be experimented with.
Trustworthy data analysis should be supported by unit testing and documentation, and the quality of this support should be one of the considerations in evaluation of the data analysis.
In other words, the article presenting the research should be part of a compendium \cite{gentleman07}.
The compendium for this article is at \url{https://github.com:jifanli/metapop_article}, with software provided as an R package.

@@

{\it Data}.
COVID-19 case reports, city population counts, and the time-varying matrix of movement between cities, were taken from \citep{li20}.
Some erroneous numbers, revealed by our data analysis, were subsequently modified as described in Supplementary Sec.~\suppSecOurModel.

{\it Model}.
Our model is a spatiotemporal partially observed Markov process (SpatPOMP) \citep{ionides21}.
A partially observed Markov process (POMP) \citep{breto09,king16} is a stochastic dynamic model comprised of: (i) a latent process having the Markov property that the future is conditionally independent of the past given the present; (ii) a measurement process, describing how the data are modeled as noisy and incomplete observations of the latent process.
The SpatPOMP framework adds an extra assumption that the latent process is comprised of a collection of units, each of which has its own latent process and observation process.
The latent processes for each unit can be interdependent, but the observations for a given unit are required to depend only on the latent process for that unit.
In our context, each city is a unit.
General notation for SpatPOMP models is set up in Supplementary Sec.~\suppSecOurModel, and this notation is subsequently used to define mathematically the specific SpatPOMP models studied in this paper (\LiMobility, \LiParams,\RevisedModelUnconstrained and \RevisedModelConstrained).
All the models under consideration have an SEAIR structure, as described in Figure~\ref{fig:flow_diagram}.

{\it Likelihood evaluation and maximization}.

{\it Model criticism}.

{\it Software environment}.
Numerical work was carried out in R \cite{R}. 
Models and data analysis methodology were developed in an R package, \code{metapoppkg} \cite{metapoppkg} which is additionally designed to assist reproducibility and extendability of our results.
Models in \code{metapoppkg} are implemented using \code{spatPomp} \cite{asfaw23arxiv} which provides a general representation of SpatPOMP models extending the POMP model representation in \code{pomp} \cite{king16}.

\section*{Data availability}

\section*{Code availability}

>>>>

\section*{Introduction}
\label{sec:intro}

@@

\section*{Review of metapopulation models and inference methods}
\label{sec:review}

@@

Population dynamics may be approximately linear on a log scale, and this has been used to develop linearization methods for epidemiological time series analysis \cite{bjornstad02} that have been extended to metapopulation analysis \cite{xia04}.
This provides a numerically convenient set of tools, but requires scientists to work within a limited class of models. The ensemble Kalman filter (EnKF) developed in the context of massive geophysical model, combines an ensemble representation with a computationally efficient update rule inspired by the scalable linear Kalman filter, providing an approach with excellent scalability \cite{evensen09book,evensen22}.
For biological systems, EnKF was first demonstrated as a computationally convenient tool for compartment models at a single location \cite{shaman12,yang14}.
Subsequently, it has been applied for epidemiological metapopulation inference \cite{li20,kramer20-plos-cb}.
However, the linearization in the EnKF filter update rule can be problematic for highly nonlinear systems \cite{evensen22}.
Further, a linear update rule is not appropriate for small, discrete populations unless EnKF is embedded within a MCMC algorithm \cite{katzfuss19}.

@@

\section*{Metapopulation analysis of COVID-19 spread in China}
\label{sec:covid}
Models are an irreplaceable tool to inform public policy, despite delicate issues in their implementation and interpretation \cite{mccabe21,saltelli20}.
Some of the difficulties are operational, others are conceptual.
Operationally, we seek to fit complex models using statistically valid, reproducible and transparent methods.
Conceptual difficulties arise when drawing causal conclusions from fitting models to observational data, giving rise to opportunities for incorrect conclusions due to missing variables or other forms of model misspecification.
A strength of metapopulation modeling is that the model can be build upon established scientific knowledge, which may give some protection against gross forms of model misspecification (for example, asserting the existence of latent dynamic variables which simply do not exist).
A model assimilated to data guarantees that assumptions have been framed in a way consistent with certain facts, and evidence for predictive skill can support the value of the model construction.
Indeed, it can be practically impossible to make sense of the nonlinear stochastic interactions driving biological dynamics without representing them via a model \cite{mccabe21,lofgren14}.
We therefore seek to harness the power of models while assessing, acknowledging and minimizing their weaknesses.

As an example, we reconsider the influential analysis of COVID-19 from early in the pandemic by Li et al. \cite{li20}, that provided estimates of transmission parameters and the effect of lockdown using the limited data available at the time.
Other teams have fitted models to address similar questions \cite{kraemer20,yang21,brett23} but the study by \cite{li20} is distinctive for fitting a stochastic mechanistic metapopulation model to extensive spatiotemporal data for 373 Chinese cities.
The results were published in May, 2020, based on reported cases from January 10 to February 8 of that year.
The state-of-the-art spatiotemporal analysis was possible on an urgent timescale because the team of researchers had developed their methodology in a sequence of previous situations \cite{shaman12,yang14,yang15,pei18}.
The paper is written with attention to reproducibility, and the main results are strengthened by various supporting analyses in an extensive supplement.
While examining the points mentioned above, we have identified various weaknesses that could have been mitigated by adhering to the aforementioned recommendations.
Our goal is not to criticize any specific paper, but rather to build on the timely analysis of \cite{li20} to demonstrate how recently developed techniques provide possibilities to carry out superior data analysis in future.

@@

We avoided the issue of model {\LiMobility} by adding some additional mobility based on a gravity movement model, as described in Supplementary Sec.~\suppSecMobility, giving rise to model~{\LiParams}.
We implemented an additional adjustment between models {\LiMobility} and {\LiParams} to align the measurement model with the ensemble Kalman filter (EnKF) inference method presented by \cite{li20}.
That EnKF implementation involved specifying a quanitity called the observation error variance, defined as a function of the observed cases, to quantify the uncertainty in the measurements.
Within the POMP specification, the measurement variance can depend on the latent state but not directly on the observed data.
To interpret the choice of EnKF observation variance within the POMP framework, we specified the measurement model for~{\LiParams} to have equivalent scaling to the choice of \citep{li20}, but with dependence on the reported cases replaced by dependence on the modeled, but unobserved, exact case count.

@@

All parameter estimates are reported in Supplementary Table~\suppTableResults.
Here, we discuss the estimated basic reproductive number (i.e., the expected number of secondary infections from one index case in a fully susceptible population), denoted by $\Rzero^{\before}$ and $\Rzero^{\after}$ before and after the January 23rd lockdown.
Our estimates for model {\RevisedModelConstrained}, are $\Rzero^{\before}=\Sexpr{myround(ci_c["R0_be", "mle"],2)}$ with confidence interval (CI) (\Sexpr{myround(ci_c["R0_be", "ci1"],2)},\Sexpr{myround(ci_c["R0_be", "ci2"],2)}) and $\Rzero^{\after}=\Sexpr{myround(ci_c["R0_af","mle"],2)}$  with CI (\Sexpr{myround(ci_c["R0_af", "ci1"],2)},\Sexpr{myround(ci_c["R0_af","ci2"],2)}), where the estimates and their associated 95\% CIs obtained by profile likelihood (Supplementary Sec.~\suppSecInference).
This implies that the Chinese government non-pharmaceutical interventions instituted on and around January 23 reduced $\Rzero$ by a factor of $\Sexpr{myround(ci_c["R0_be","mle"]/ci_c["R0_af","mle"],1)}$.
By contrast, the estimates of \cite{li20} are $\widetilde\Rzero^{\before}=2.38$  with CI $(2.03,2.77)$ and $\widetilde\Rzero^{\after}=0.98$  with CI $(0.83,1.16)$, implying reduction by a factor of $\Sexpr{myround(2.38/0.98,1)}$.
For comparison, interventions implemented across a panel of 41 countries (34 European) were estimated to reduce $\Rzero$ by a factor of $\Sexpr{myround(1/(1-0.77),1)}$  with CI $(\Sexpr{myround(1/(1-0.65),1)},\Sexpr{myround(1/(1-0.85),1)})$ \citep{brauner21}.
Our estimate for $\Rzero$ before lockdown is toward the high end of comparable estimates, reviewed by \citep{park20-jcm}.
An alternative metaopopulation analysis of the pre-lockdown China data, with a deterministic transmission model, obtained an $\Rzero$ estimate of 3.11  with CI  (2.39,4.13) \citep{read21}.

Our model inherits the property of \citep{li20} that infections arising during the pre-lockdown period will generally be reported during the lockdown, due to the reporting delay modeled as a distributed delay with a mean of 9 days pre-lockdown and 6 days post-lockdown.
Thus, the model is permitted to explain the data by inferring rapid, unreported spread prior to January 23.
Despite this shared constraint on the form of the model, conclusions of our analysis differ from \cite{li20}.
Beyond the estimates of $\Rzero$, a notable difference is that we find the estimated transmissibility of observed cases is close to that of unobserved cases, especially before lockdown (Supplementary Table~S1).

@@

\section*{Discussion}
\label{sec:conclusion}

@@

Software environments are critical for providing data analysis that is not only reproducible but also readily extendable.
Scientists developing a data analysis should develop an environment that empowers them to explore their own models and data, and then they should share this environment as part of the publication process.
In practice, this involves encapsulating data analysis within a software package  that immerses the user in a documented environment where the models, methods and data used for the article can be readily be experimented with.
Trustworthy data analysis should be supported by unit testing and documentation, and the quality of this support should be one of the considerations in evaluation of the data analysis.
In other words, the article presenting the research should be part of a compendium \cite{gentleman07}.
The compendium for this article is comprised of the article source code, at \url{https://github.com:jifanli/metapop_article}, together with the software environment for the data analysis, provided by the R package at \url{https://github.com:jifanli/metapoppkg}.

@@

\noindent {\it Data}.
COVID-19 case reports, city population counts, and the time-varying matrix of movement between cities, were taken from \citep{li20}.
Some erroneous numbers, revealed by our data analysis, were subsequently modified as described in Supplementary Sec.~\suppSecOurModel.

\vspace{3mm}

\noindent {\it Model}.
Our model is a spatiotemporal partially observed Markov process (SpatPOMP) \citep{ionides21}.
A partially observed Markov process (POMP) \citep{breto09,king16} is a stochastic dynamic model comprised of: (i) a latent process having the Markov property that the future is conditionally independent of the past given the present; (ii) a measurement process, describing how the data are modeled as noisy and incomplete observations of the latent process.
The SpatPOMP framework adds an extra assumption that the latent process is comprised of a collection of units, each of which has its own latent process and observation process.
The latent processes for each unit can be interdependent, but the observations for a given unit are required to depend only on the latent process for that unit.
In our context, each city is a unit.
General notation for SpatPOMP models is set up in Supplementary Sec.~\suppSecOurModel, and this notation is subsequently used to define mathematically the specific SpatPOMP models studied in this paper (\LiMobility, \LiParams,\RevisedModelUnconstrained and \RevisedModelConstrained).
All the models under consideration have an SEAIR structure, as described in Figure~\ref{fig:flow_diagram}.

\vspace{3mm}

\noindent {\it Likelihood evaluation and maximization}.
The log-likelihood for the SpatPOMP models was calculated using a block particle filter \cite{rebeschini15,asfaw23arxiv}.
This log-likelihood was then maximized using an iterated block particle filter \citep{ionides22,ning23}
Further discussion of these algorithms is in Supplementary Sec.~\suppSecIBPF.
Using this maximization procedure, we constructed confidence intervals by profile likelihood, employing Monte Carlo adjusted profiles \cite{ionides17,ning21} to correct for Monte Carlo variability.

\vspace{3mm}

\noindent {\it Model criticism}.
A negative binomial autoregressive model was used to provide a non-mechanistic benchmark log-likelihood, as described in Supplementary Sec.~\suppSecBenchmark.
This model was also used to construct benchmark conditional log-likelihoods for each separate observation.
These, differenced from the corresponding SEAIR log-likelihoods, were used to define anomalies.
The anomalies were explored to identify data points which were poorly explained by the model (Supplementary Sec.~\suppSecResiduals).
In preliminary data analysis, these anomalies helped to identify some errors in the data which were subsequently corrected (Supplementary Sec.~\suppSecLiModel).

\vspace{3mm}

\noindent {\it Software environment}.
Numerical work was carried out in R \cite{R}. 
Models and data analysis methodology were developed in an R package, \code{metapoppkg}, which is additionally designed to assist reproducibility and extendability of our results.
Models in \code{metapoppkg} are implemented using \code{spatPomp} \cite{asfaw23arxiv} which provides a general representation of SpatPOMP models extending the POMP model representation in \code{pomp} \cite{king16}.

\section*{Data availability}
The data are available as part of the \code{metapoppkg} R package.

\section*{Code availability}
The code generating the manuscript is archived on Zenodo
\eic{ZENODO MS LINK HERE} and a development version is at \url{https://github.com:jifanli/metapop_article}.
The software environment developed for this analysis is also available as both an archived version (\eic{ZENODO METAPOPPKG LINK HERE}) and a development version (\url{https://github.com:jifanli/metapoppkg}).
These programs depend on other open-source software archived at \url{https://cran.r-project.org/}.
